---
ve_data_science:
  title: "Updating carbon use efficiency parameters"

  description: |
      This analysis involves a GLM to estimate parameters for
      temperature-dependent microbial carbon use efficiency (CUE)
      that does not predict CUE out of bound (stays between 0 and 1).
      It also updates the soil constant --- currently the one we use
      originates from a rice paddy study in the US, which is too
      narrow. The data used here is a global dataset. I re-analysed the
      dataset using a GLM, because the original study used a Gaussian
      regression. See the output html file for more details.

  author:
    - Hao Ran Lai

  virtual_ecosystem_module:
    - Soil

  status: final

  input_files:
    - name: 41598_2019_42145_MOESM2_ESM.csv
      path: data/primary/soil/carbon_use_efficiency/Qiao2019
      description: |
        The global carbon-use-efficiency (CUE) dataset was compiled by
        Qiao et al. (2019) https://doi.org/10.1038/s41598-019-42145-6
        They shared the data in the Supplementary ("Dataset 1")
        The dataset also contains treatment temperature that is used as a
        predictor of CUE in our analysis here.
        The original data was published as a .xlsx file, which is currently
        gitignored. I have saved it as a csv file so it can be pushed for
        code checking, for now.

  output_files:
    - name: update_cue_parameters.md
      path: analysis/soil/carbon_use_efficiency
      description: |
        This is a RMarkdown report of my analysis, including the results.
        Need to knitr from the RMarkdown file.
        If you run the script, it will also save the GLM model object as follows.
    - name: model_cue.rds
      path: analysis/soil/carbon_use_efficiency
      description: |
        A Bayesian GLM model object in the class brms, which contains the
        summary statistics, as well as the full posterior samples

  package_dependencies:
      - tidyverse
      - readxl
      - sf
      - tmap
      - brms

  usage_notes: |
    The Bayesian model takes a few minutes to run, grab a cuppa! If the setup
    works as intended, then the model will only run once unless you make changes.
    Run jupytext --to ipynb code/soil/cue/update_cue_parameters.Rmd
    --output code/soil/cue/update_cue_parameters.ipynb
    to convert this Rmd file to ipynb

output:
    github_document:
        html_preview: false
        
bibliography: ../../../bib/refs.bib

---

```{r setup, include=FALSE}
# some knit options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = NA,
  message = NA,
  fig.align = "center",
  dpi = 72
)
```

```{r Libraries}
# load packages
library(tidyverse)
library(sf)
library(tmap)
library(brms)
library(readxl)
```

# Background

```{r Read data, message=FALSE, warning=FALSE}
# inout folder
in_path <-
  "data/primary/soil/carbon_use_efficiency/"

dat_cue <-
  read_xlsx(
    paste0(
      in_path,
      "41598_2019_42145_MOESM2_ESM.xlsx"
    )
  ) %>%
  select(No_Wj,
    Author,
    # Site,
    lat = Latitude,
    lon = longitude,
    CUE,
    PFT = `PFT...7`,
    Temp = "Temperature(℃)"
  ) %>%
  # fill in and simplify author names
  fill(Author, .direction = "down") %>%
  mutate(AuthorID = as.numeric(as.factor(Author))) %>%
  # center temperature at 20 deg C
  mutate(
    Temp = as.numeric(Temp),
    Temp_centered = Temp - 20
  ) %>%
  # remove incomplete data
  filter(!is.na(Temp_centered))
```

Currently, the temperature-dependent carbon use efficiency (CUE) function we used is

$$
\text{CUE} = \text{CUE}_\text{ref} + \text{CUE}_T \left(T - T_\text{ae-ref} \right) \,,
$$

where $T$ is temperature ($^\circ$C), $\text{CUE}_\text{ref}$ is the CUE at reference
temperature $T_\text{ae-ref}$ (i.e., centered intercept), and $\text{CUE}_T$ is the
temperature-dependence slope. This equation came from @abramoff_millennial_2018, who
cited @xu-ri_terrestrial_2008, who then cited @Devevre2000 and used their empirical
estimates. In @Devevre2000, CUE came from a single rice paddy study:

> by soil microorganisms during rice straw decomposition was determined in a rice paddy
> soil, under aerobic and anaerobic (flooded) conditions at different temperatures
> (5, 15, and 25°C)

The latest global study I could find is @Qiao2019. Mostly from US and Europe and not a
lot of tropical studies (see map below; there are some misplaced coordinates but I think
they are placeholder for missing data), but seems to be the most extensive to date.
Their model is also identical to ours so should be easy to adapt.

```{r Map, fig.height=3.5, fig.width=6}
data("World")

dat_sf <-
  dat_cue %>%
  st_as_sf(
    coords = c("lon", "lat"),
    crs = st_crs(World)
  )

ggplot() +
  geom_sf(data = World) +
  geom_sf(data = dat_sf) +
  theme_minimal()
```

@Qiao2019 performed curve fitting and published their parameter estimates. However, they
fitted a linear regression that could predict CUE out of bound (above 1 or below 0)
under certain temperature. Our goal here is to refit a model on their compiled dataset,
and constrain CUE to be bound between 0 and 1 using a generalised linear regression
(GLM).

## Modelling

A beta-distribution GLM with logit link would do this. Our generative model is:

\begin{equation*}
\begin{aligned}
\text{CUE}_{ij} &\sim \text{Beta}\left( \mu_{ij}, \phi \right) \\
\text{logit}\left( \mu_{ij} \right) &= \text{CUE}_\text{ref}
    + \text{CUE}_T \left(T_{ij} - T_\text{ae-ref} \right) + \epsilon_j \,,
\end{aligned}
\end{equation*}

where $\mu_{ij}$ is the linear predictor of CUE for observation $i$ in study $j$. The
dataset contains multiple CUE values from the same study, so I controlled for
non-independence within study using a random intercept term $\epsilon_j$.

There are covariates that could potentially be included, such as biome and length of
incubation. I have left them out for simplicity here, and because biome contains missing
data. If needed, we could revisit these covariates.

The model was fitted with Bayesian inference using the `brms` package in `R`.

```{r Model}
model <- brm(
  CUE ~ 1 + Temp_centered + (1 | Author),
  data = dat_cue,
  family = Beta(),
  prior =
    prior(normal(0, 0.5), ub = 0, class = b),
  warmup = 3000,
  iter = 4000,
  cores = 4,
  file = "analysis/soil/carbon_use_efficiency/model_cue",
  file_refit = "on_change"
)
```

## Results

```{r Parameter estimates, include=FALSE}
param <- fixef(model)
```

The posterior median of the parameters are:

- $\text{CUE}_T =$ `r round(param["Temp_centered", "Estimate"], 3)`
- $\text{CUE}_\text{ref} =$ `r round(param["Intercept", "Estimate"], 3)`

Note that the intercept $\text{CUE}_\text{ref}$ is in logit scale, in the original
proportional scale it is `r round(plogis(param["Intercept", "Estimate"]), 3)`.

```{r Prediction}
#| fig.height: 5
#| fig.width: 7
#| fig.cap: >
#|   Fitted regressions through the Qiao et al. (2019) dataset. Line color denotes the
#|   three different models: black = the Abramoff Millenial model current used by VE;
#|   red = the fitted linear regression by Qiao et al. (2019); blue = our GLM fit.
#|   Dashed blue lines are the 90% credible intervals of our GLM.

# counterfactual data
newdat <-
  data.frame(
    Temp = seq(-10, 50, length.out = 100),
    Author = NA
  ) %>%
  mutate(Temp_centered = Temp - 20)

# fitted line from our model
cue_hat <-
  fitted(
    model,
    newdata = newdat,
    scale = "response",
    robust = TRUE,
    probs = c(0.05, 0.95),
    allow_new_levels = TRUE
  )

# fitted line from Qiao et al. (2019)
cue_hat_qiao <- 0.475 - 0.016 * newdat$Temp_centered

# fitted line from Abramoff et al. (2018)
cue_hat_abramoff <- 0.6 - 0.012 * (newdat$Temp - 15)

# plot
par(mar = c(4, 4, 0.5, 0.5), las = 1)
with(dat_cue, plot(
  jitter(Temp, 2), CUE,
  xlim = c(-10, 50),
  xlab = "Temperature (\u00B0C)",
  ylab = "CUE (unitless)"
))
abline(h = c(0, 1), col = "darkgrey", lty = 2)
lines(newdat$Temp, cue_hat_qiao,
  col = "tomato",
  lwd = 2
)
lines(newdat$Temp, cue_hat_abramoff,
  col = "black",
  lwd = 2
)
lines(newdat$Temp, cue_hat[, "Estimate"],
  col = "steelblue2",
  lwd = 2
)
lines(newdat$Temp, cue_hat[, "Q5"],
  col = "steelblue2",
  lwd = 2, lty = 2
)
lines(newdat$Temp, cue_hat[, "Q95"],
  col = "steelblue2",
  lwd = 2, lty = 2
)
legend("topright",
  lwd = 2,
  col = c("tomato", "black", "steelblue2"),
  legend = c("Qiao", "Abramoff", "Our model")
)
```

# References
