---
title: "Sensitivity analysis of soil and litter input data"
author:
  name: Lai, Hao Ran
date: last-modified
format:
  html:
    comments:
      hypothesis: true
knitr:
  opts_chunk:
    echo: false
    message: false
    dpi: 300
abstract: Placeholder
---

```{r}
#| label: load-libraries
library(tidyverse)
library(sensobol)
library(tidync)
library(arrow)
library(here)
```


```{r}
#| label: crop-hook
# create a hook named crop that is associated with the knitr::hook_pdfcrop()
# function so we can use crop: true in a code chunk to call it and crop the
# displayed graph.
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
```
# Introduction


# Methods


# Results & Discussion



```{r}
#| label: read-sobol-matrix
sobol_mat <- read_rds(
  here("data/scenarios/sensitivity_soil_litter/data/sobol_matrix.rds")
)
pars <- colnames(sobol_mat)
n_samples <- nrow(sobol_mat) / (length(pars) + 2)
```



```{r}
#| label: read-merged-data
#| cache: true
merged <-
  open_dataset(
    here(
      "data/scenarios/sensitivity_soil_litter/out/all_continuous_data_merged.parquet"
    )
  ) |>
  filter(time_index == 24) %>%
  collect()
```



```{r}
#| label: calc-sobol-indices
#| cache: true
#| dependson: "merged"
Y_vars <- unique(merged$variable) |> setdiff(c("timestamp"))
forcings <-
  sens <- map(
    Y_vars,
    \(Y_var) {
      Y <-
        merged |>
        filter(variable == Y_var) |>
        mutate(scenario = as.numeric(scenario)) |>
        right_join(
          data.frame(scenario = seq_len(nrow(sobol_mat))),
          by = join_by(scenario)
        ) |>
        arrange(scenario) |>
        pull(value)
      sobol_indices(Y = Y, N = n_samples, params = pars) |>
        pluck("results") |>
        as.data.frame() |>
        mutate(Y_var = Y_var)
    }
  ) |>
    list_rbind() |>
    group_by(Y_var) |>
    # remove in-development or problematic output variables with NA values or
    # zero variation
    filter(!is.na(original), var(original) != 0) |>
    # "normalise" sobol indices for comparative visualisation
    # mutate(original_s = original / sd(original)) |>
    ungroup()
```

Overall sensitivity analysis results

```{r}
#| label: fig-overall
#| dependson: "sens"
#| fig-cap: "Placeholder."
#| column: screen
#| crop: true
sens_total <- sens |> filter(sensitivity == "Ti")
sens_total |>
  ggplot() +
  geom_raster(aes(y = parameters, x = Y_var, fill = original)) +
  scale_fill_viridis_c(
    option = "cividis",
    transform = "sqrt",
    name = "Sensitivity"
  ) +
  labs(x = "Output variables", y = "Input variables") +
  coord_fixed(expand = FALSE) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 3, angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 3),
    legend.position = "top"
  )
```

Compare most and least sensitive pairs

```{r}
sens_min_max <-
  sens_total |>
  filter(original == min(original) | original == max(original))

foo <-
  open_dataset(
    here(
      "data/scenarios/sensitivity_soil_litter/out/all_continuous_data_merged.parquet"
    )
  ) |>
  filter(
    time_index == 24,
    variable %in%
      c("litter_pool_above_metabolic_cnp_C", "animal_bacteria_consumption")
  ) |>
  collect() |>
  pivot_wider(names_from = "variable", values_from = "value")
```

Experimental phase diagram

```{r}
with(foo, plot(litter_pool_above_metabolic_cnp_C, animal_bacteria_consumption))
```


Look at histogram


```{r}
sens |>
  filter(sensitivity == "Ti") |>
  ggplot() +
  geom_histogram(aes(original))
```

Which factor is dominating this uncertainty?

Are there a few dominating factors or is the uncertainty originating from several factors?

Do the factors act singularly or in combination with one another?


# Reference


# Appendix
