---
title: Empty forest? We need a scenario with persistent animal populations over the simulation period
author:
  name: Lai, Hao Ran
date: last-modified
format:
  html:
    minimal: true
    comments:
      hypothesis: true
  gfm: default
  ipynb:
    execute:
      eval: true
      warning: false
      enabled: true
  pdf: default
  docx: default
toc: true
theme: sandstone
knitr:
  opts_chunk:
    message: false
code-fold: true
---

# Preamble

I am conducting a sensitivity analysis for the soil and litter modules. A sensitivity analysis examines how much of the variation in an output is attributed to variation in an input. **However, if an output never varies, it is meaningless to conduct a sensitivity analysis.** This happens to a few animal-related outputs in the `all_continuous_data.nc` file. My gut feeling is that the lack of temporal variation is due to the animal FGs dying off, hence the exploration here.

At the end of this report, I explain why we might want to design a scenario where there is at least some persistent animal populations, at least for the purpose of sensitivity analyses.

# Model and data summary

I ran the full `maliau_2` scenario:

- config in `data/scenarios/maliau/maliau_2/config`
- data in `data/scenarios/maliau/maliau_2/data`
- VE version: v0.1.1a17 (dev version; commit [b6a216](https://github.com/ImperialCollegeLondon/virtual_ecosystem/commit/b6a216bd263a50adfb5e1c90eb64583699578017))
- OS: Windows 11

::: {.callout-note}
# Improvement
I'm thinking of writing a function to automatically summarise key model configurations from the full config TOML compiled by `ve_run`. We know what `maliau_2` means but others outside our team may not remember. Sometimes a scenario config may also change. Tracked in [this Issue](https://github.com/ImperialCollegeLondon/ve_data_science/issues/327).
:::

# Animal continuous state variables

Currently, I'm examining:

- `animal_arbuscular_mycorrhiza_consumption`
- `animal_bacteria_consumption`
- `animal_ectomycorrhiza_consumption`
- `animal_pom_consumption_cnp`
- `animal_saprotrophic_fungi_consumption`
- `total_animal_respiration`

::: {.callout-caution}
# Need help
Pardon if I haven't grasped the full list of "animal-related" continuous state variables. Please suggest anything that is missing here.
:::

```{r}
#| label: load-packages
library(tidync)
library(tidyverse)
library(here)
source(here("tools/R/tidy_continuous_data.R"))
```

```{r}
#| label: get-data
animal_vars <- c(
  "animal_arbuscular_mycorrhiza_consumption",
  "animal_bacteria_consumption",
  "animal_ectomycorrhiza_consumption",
  "animal_pom_consumption_cnp",
  "animal_saprotrophic_fungi_consumption",
  "total_animal_respiration"
)
animal_cont <- tidy_continuous_data(
  here("data/scenarios/maliau/maliau_2/out/all_continuous_data.nc"),
  variables = animal_vars
)
```

First I saw that the range of these state variables are very small. Are they truly very small, or are they numerical imprecisions that need to be zapped to zero?
```{r}
#| label: summary-table
animal_cont |>
  group_by(variable) |>
  summarise(min = min(value), max = max(value))
```

Here's how the variables looked over simulation time steps:
```{r}
#| label: fig-temporal-trend
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "Temporal trends in animal state variables. Each semi-transparent line is a grid cell."
animal_cont |>
  unite("variable2", variable, element, na.rm = TRUE) |>
  ggplot() +
  facet_wrap(~variable2, ncol = 1, scales = "free_y") +
  geom_line(aes(time_index, value, group = cell_id), alpha = 0.5) +
  theme_bw()
```

```{r}
#| label: last-cohort
animal_cohort <- read_csv(
  here("data/scenarios/maliau/maliau_2/out/animal_cohort_data.csv")
)
max_cohort_time <- max(animal_cohort$time_index)
```

Before proceeding, I checked the animal cohort data and saw that all cohorts went extinct after time step `r max_cohort_time`.

Following Nick's suggestion, I also checked the temporal trends in resource availability:
```{r}
#| label: fig-resource-trend
#| fig-height: 6
#| fig-width: 7
#| fig-cap: "Temporal trends in resource state variables. Each semi-transparent line is a grid cell."
resource_vars <- c(
  "soil_c_pool_arbuscular_mycorrhiza",
  "soil_c_pool_bacteria",
  "soil_c_pool_ectomycorrhiza",
  "soil_c_pool_saprotrophic_fungi",
  "soil_cnp_pool_pom"
)

resource_cont <- tidy_continuous_data(
  here("data/scenarios/maliau/maliau_2/out/all_continuous_data.nc"),
  variables = resource_vars
)

resource_cont |>
  unite("variable2", variable, element, na.rm = TRUE) |>
  ggplot() +
  facet_wrap(~variable2, ncol = 1, scales = "free_y") +
  geom_line(aes(time_index, value, group = cell_id), alpha = 0.5) +
  theme_bw()
```

A few follow-up questions upon seeing the temporal graphs:

- Why do we still see non-zero values in some variables long after all animals have gone extinct since time step `r max_cohort_time`?
- Presumably these variables are positive only; what do the negative values mean? The way they fluctuate almost symmetrically around zero makes me suspect that the non-zero values are not true non-zeros but numerical imprecision.
- There seems to be some relationship with resource availability. But there is no animal to consume then at later time steps?

::: {.callout-caution}
# Checkpoint
If these trends are numerical artefacts rather than true consumption and respiration rates, then there is not much point to read on.
:::

# Why we need persistent animal populations

Mainly so that we can include animal-related state variables into the sensitivity analyses. More importantly, the animal variables feed back into the non-animal variables. Unless we are truly aiming for an empty-forest scenario, we will be left with a half-complete sensitivity analysis.

Should we consider an alternative set of animal FG definitions? Currently `maliau_2` uses the level 1 definition, which contain only a single herbivorous endotherm that always go extinct very early on. Has anyone run VE with the level 2 definitions? If the level 2 groups also go extinct, should we consider an alternative set (perhaps more basal in tropic levels) that can persist over time, and hence continue to keep the animal and non-animal components coupled until the end of simulation?
