---
title: "SAFE database screening for validation datasets"
author:
  name: Lai, Hao Ran
date: last-modified
format:
  html:
    embed-resources: true
comments:
  hypothesis: true
toc: true
reference-location: margin
code-fold: true
knitr:
  opts_chunk:
    message: false
---

# Goal

To record my decisions as I screen through the SAFE database[^1] on Zenodo for soil and litter validation datasets.

[^1]: I'm using the SAFE database as a testing example before expanding the workflow here to all other datasets in the wild.

I also want something sharable to discuss with Jacob and the team. I expect a few rounds of discussion that iterate the decisions here.

I expected most of the datasets to be not suitable, but I am including everything that I've screened through here so I remember what has been inspected and why it has been included or excluded.

# Methods

I wanted the final product to be a table of decision per dataset, plus the dataset' metadata (mostly citation information). But I do not want to manually copy-and-paste into MS Excel, so I opted for a custom function `log_dataset()` in R, which (1) automates the metadata extraction and (2) records my decision into a YAML file for each dataset. All YAMLs are then combined into @tbl-tabulate-records.[^2]

[^2]: An unexpected feeling from generating a table here as a report (compared to an Excel sheet) is that I could add paragraphs of background text and thoughts in the same document, rather than having to write an email or GitHub issue and attach the Excel sheet, which are hard to track.

```{r}
#| label: load-libraries
library(here)
library(yaml)
library(DT)
library(tidyverse)
```

```{r}
#| label: read-yaml
# list all the dataset logs
logdir <- here("data/derived/soil/validation/config/sources")
logfiles <- list.files(logdir, pattern = "\\.yaml$")

# read the logs and tidy up
records <-
  file.path(logdir, logfiles) |>
  map(read_yaml) |>
  map2(logfiles, \(x, nm) as.data.frame(x) |> mutate(logfile = nm)) |>
  list_rbind()
```

# Notes for discussion

- I will next focus on datasets that are listed as "included" under the "Decision" column. There are currently `r filter(records, decision == "included") |> nrow()` datasets.
- Among the excluded datasets, a lot of them contains soil temperature and soil moisture, which fall under the abiotic and hydrology modules, but I will discuss with Lela if I should be responsible for them.
- If we are really low on validation datasets, then datasets that are `used_elsewhere` (see Decision column) for input data can be split into testing--training sets. However this entails quite a lot of change to the previous input-data codes.
- As expected, some of the datasets here that are not useful for soil can be useful for other modules, but I haven't decided on a good way to document and communicate that.

```{r}
#| label: tbl-tabulate-records
#| column: screen
records |>
  mutate(
    DOI = paste0("<a href='", metadata.url, "' target='_blank'>", doi, "</a>")
  ) |>
  select(
    DOI,
    Log = logfile,
    Title = metadata.title,
    Decision = decision,
    Reason = reason,
    Notes = notes,
    Year = metadata.year
  ) |>
  datatable(
    caption = "Screening decision of the SAFE database for soil and litter validation datasets.",
    filter = list(position = "top"),
    options = list(pageLength = 25),
    escape = FALSE
  )
```
