---
title: "SAFE database screening for validation datasets"
author:
  name: Lai, Hao Ran
date: last-modified
format:
  html: 
    embed-resources: true
comments:
  hypothesis: true
toc: true
reference-location: margin
code-fold: true
knitr:
  opts_chunk:
    message: false
---

# Goal

To record my decisions as I screen through the SAFE database[^1] on Zenodo for soil and litter validation datasets.

[^1]: I'm using the SAFE database as a testing example before expanding the workflow here to all other datasets in the wild.

I also want something sharable to discuss with Jacob and the team. I expect a few rounds of discussion that iterate the decisions here.

I expected most of the datasets to be not suitable, but I am including everything that I've screened through here so I remember what has been inspected and why it has been included or excluded.

# Methods

I wanted the final product to be a table of decision per dataset, plus the dataset' metadata (mostly citation information). But I do not want to manually copy-and-paste into MS Excel, so I opted for a custom function `log_dataset()` in R, which (1) automates the metadata extraction and (2) records my decision into a YAML file for each dataset. All YAMLs are then combined into @tbl-tabulate-records.

```{r}
#| label: load-libraries
library(here)
library(yaml)
library(DT)
library(tidyverse)
```

```{r}
#| label: read-yaml
# list all the dataset logs
logfiles <- list.files(
  here("data/derived/soil/validation/records"),
  pattern = "\\.yaml$",
  full.names = TRUE
)

# read the logs and tidy up
records <-
  logfiles |>
  map(read_yaml) |>
  map(as.data.frame) |>
  list_rbind()
```

# Notes for discussion

- I will next focus on datasets that are listed as "included" under the "Decision" column. There are currently `r filter(records, decision == "included") |> nrow()` datasets.
- Among the excluded datasets, a lot of them contains soil temperature and soil moisture, which fall under the abiotic and hydrology modules, but I will discuss with Lela if I should be responsible for them.
- If we are really low on validation datasets, then datasets that are `used_elsewhere` (see Decision column) for input data can be split into testing--training sets. However this entails quite a lot of change to the previous input-data codes.
- As expected, some of the datasets here that are not useful for soil can be useful for other modules, but I haven't decided on a good way to document and communicate that.

```{r}
#| label: tbl-tabulate-records
#| column: screen
records |>
  select(
    DOI = doi,
    Title = metadata.title,
    Decision = decision,
    Reason = reason,
    Notes = notes,
    Year = metadata.year
  ) |>
  datatable(
    caption = "Screening decision of the SAFE database for soil and litter validation datasets.",
    filter = list(position = "top"),
    options = list(pageLength = 25)
  )
```