---
title: "Soil and Litter GitHub Issues Tracker"
author:
  name: Lai, Hao Ran
date: last-modified
format:
  html:
    toc: true
    toc-expand: true
    theme: sandstone
    embed-resources: true
    other-links:
      - text: "GitHub Issues"
        icon: github
        href: https://github.com/ImperialCollegeLondon/ve_data_science/issues?q=is%3Aissue%20state%3Aopen%20author%3Ahrlai
      - text: "GitHub Milestones"
        icon: github
        href: https://github.com/ImperialCollegeLondon/ve_data_science/milestones
output-file: "VE_Task Breakdown Soil Litter Hao Ran.html"
title-block-banner: true
comments:
  hypothesis: true
execute:
  echo: false
  warning: false
abstract-title: Preamble
abstract: |
    This is a semi-automated report that I update weekly. 
    The [Overview](#sec-overview) section is written manually, while the 
    [Subtasks](#sec-milestones) section is synchronised automatically with 
    [GitHub Issues](https://github.com/ImperialCollegeLondon/ve_data_science/issues?q=is%3Aissue%20state%3Aopen%20author%3Ahrlai) and 
    [Milestones](https://github.com/ImperialCollegeLondon/ve_data_science/milestones). 
    This report should reflect most of my tasks, which are almost 100% tracked
    on GitHub.
    The [Subtasks](#sec-milestones) GitHub Issue tracker tables show only Open
    issues by default. To view Closed issues, simply change the column's
    filter.
---

```{r}
#| label: setup
library(DT)
library(tidyverse)
library(IssueTrackeR)
library(htmltools)

# IssueTrackerR configs
options(IssueTrackeR.dataset.dir = "data/derived/soil/gh_issues")
options(IssueTrackeR.owner = "ImperialCollegeLondon")
options(IssueTrackeR.repo = "ve_data_science")

# Set global options for all DT tables
options(
  DT.options = list(
    pageLength = 5,
    autoWidth = TRUE,
    # 2. Pre-populate specific column filters
    searchCols = list(
      NULL,
      NULL,
      NULL,
      list(search = '["open"]'),
      NULL
    )
  )
)

# function to render the interactive table
render_issues <- function(issue) {
  issue |>
    select(-milestone) |>
    datatable(
      rownames = FALSE,
      colnames = c(
        "Issue" = "number",
        "Title" = "title",
        "Body (truncated)" = "body",
        "State" = "state",
        "Created at" = "created_at"
      ),
      filter = "top",
      escape = 2
    )
}

# function for a HTML progress bar
progress_bar <- function(x, total = 100, color = "#2c7bb6") {
  pct <- round(x / total * 100)
  lbl <- paste0(pct, "% complete; ", total - x, " open; ", x, " closed")

  tags$div(
    style = "margin: 1rem 0",
    tags$div(
      style = "display: flex; justify-content: space-between; font-size: 0.85rem; color: #555; margin-bottom: 4px;",
      tags$span("Progress"),
      tags$span(lbl)
    ),
    tags$div(
      style = "background: #e9ecef; border-radius: 999px; height: 10px; overflow: hidden;",
      tags$div(
        style = glue::glue(
          "width: {pct}%; height: 100%; background: {color};
           border-radius: 999px; transition: width 0.6s ease;"
        )
      )
    )
  )
}

# function that applies progress_bar() to a milestone object
progress_bar_milestone <- function(milestone, ...) {
  tmp <- milestones |> filter(title == milestone)
  n_closed <- tmp$nb_issues_closed
  n_total <- tmp$nb_issues_open + tmp$nb_issues_closed
  progress_bar(n_closed, n_total, ...)
}
```

```{r}
#| label: read-issues
# download GitHub issues
issues <- get_issues(source = "online", state = "all", verbose = FALSE)
milestones <- get_milestones(source = "online", state = "all", verbose = FALSE)

# write_to_dataset(issues)
# update_database()

# prepare the issue tracker table
issue_table <-
  issues |>
  as_tibble() |>
  filter(creator == "hrlai") |>
  # Convert NA milestones to a string for grouping
  mutate(milestone = replace_na(as.character(milestone), "No Milestone")) |>
  # Arrange by milestone, then by our custom state order
  mutate(state = factor(state, levels = c("open", "closed"))) |>
  arrange(milestone, state, desc(created_at)) |>
  # cleaner dates
  mutate_at(vars(created_at, closed_at), ymd) |>
  # Create a column for the linked number
  mutate(
    number = sprintf('<a href="%s" target="_blank">%s</a>', html_url, number),
    body = str_trunc(body, 100),
  ) |>
  select(number, title, body, state, milestone, created_at)
```

# Overview {#sec-overview}

## Lastest description

I am finalising two main things to bring over to Silwood for in-person discussions:

- Validation database and dataset screening pipeline for soil and litter; lots of discussion space around unit conversion and harmonisation
- A draft sensitivity analysis pipeline for soil and litter; related to HPC use

## Required elements ("Waiting for")

- Urgent
  - Code review for Set up soil and litter validation screening and database (PR [#339](https://github.com/ImperialCollegeLondon/ve_data_science/pull/339))
  - Taran to review 329 examine animal continuous state variables from maliau_2 (PR [#330](https://github.com/ImperialCollegeLondon/ve_data_science/pull/330))
  - VE Issue: Error: Trophic transfer masses must be non-negative [#1628](https://github.com/ImperialCollegeLondon/virtual_ecosystem/issues/1628)
- Not urgent
  - HPC tutorial: I am able to submit jobs, but want to double check some best practices
  - Team to settle on reporting format
  - Team to settle on folder structure
  - A bug-free VE release that runs on Maliau 1 and 2 scenarios

## Additional notes

- None currently.

# Subtasks (as GitHub Milestones) {#sec-milestones}

## Soil parameterisation

Parameterise soil and litter constants with empirical data or best-guess estimates.

```{r}
#| label: parameterisation
# milestone overview
progress_bar_milestone("Soil parameterisation")

# issue table
issue_table |>
  filter(milestone == "Soil parameterisation") |>
  render_issues()
```

## Soil validation

Validate soil and litter variables.

```{r}
#| label: validation
# milestone overview
progress_bar_milestone("Soil validation")

# issue table
issue_table |>
  filter(milestone == "Soil validation") |>
  render_issues()
```

## Soil calibration

Not started yet.

## Soil sensitivity

Sensitivity analysis for soil and litter.

```{r}
#| label: sensitivity
# milestone overview
progress_bar_milestone("Soil sensitivity")

# issue table
issue_table |>
  filter(milestone == "Soil sensitivity") |>
  render_issues()
```

## Soil input data (phase 2)

Follow up to polish soil and litter input data for completeness.

```{r}
#| label: input2
# milestone overview
progress_bar_milestone("Soil input data (phase 2)")

# issue table
issue_table |>
  filter(milestone == "Soil input data (phase 2)") |>
  render_issues()
```

## Soil input data (phase 1)

MVP of soil and litter input datasets using "Maliau" datasets. This milestone has been fully completed.

```{r}
#| label: input1
# milestone overview
progress_bar_milestone("Soil input data (phase 1)")

# issue table
issue_table |>
  filter(milestone == "Soil input data (phase 1)") |>
  render_issues()
```

## Other tasks

These are miscellaneous tasks not targetting any milestones, but are important for housekeeping and repo maintenance.

```{r}
#| label: other-tasks
# issue table
issue_table |>
  filter(milestone == "No Milestone") |>
  render_issues()
```